# Prerequisite
## Load Data
```{r}
df = read.csv("StudentPerformanceFactors.csv")
head(df)
```

## Libraries
```{r}
options(repos = c(CRAN = "https://cloud.r-project.org"))
if (!require(dplyr)) install.packages("dplyr")
if (!require(ggplot2)) install.packages("ggplot2")
if (!require(lmtest)) install.packages("lmtest")
if (!require(nortest)) install.packages("nortest")

library(dplyr)
library(ggplot2)
library(lmtest)
library(nortest)
```

# 3
## Encode Qualitative Variables
```{r}
df <- df %>%
  mutate(
    Parental_Involvement = case_when(
      Parental_Involvement == "Low" ~ 1,
      Parental_Involvement == "Medium" ~ 2,
      Parental_Involvement == "High" ~ 3
    ),
    Access_to_Resources = case_when(
      Access_to_Resources == "Low" ~ 1,
      Access_to_Resources == "Medium" ~ 2,
      Access_to_Resources == "High" ~ 3
    ),
    Motivation_Level = case_when(
      Motivation_Level == "Low" ~ 1,
      Motivation_Level == "Medium" ~ 2,
      Motivation_Level == "High" ~ 3
    ),
    Family_Income = case_when(
      Family_Income == "Low" ~ 1,
      Family_Income == "Medium" ~ 2,
      Family_Income == "High" ~ 3
    ),
    Teacher_Quality = case_when(
      Teacher_Quality == "Low" ~ 1,
      Teacher_Quality == "Medium" ~ 2,
      Teacher_Quality == "High" ~ 3
    ),
    Distance_from_Home = case_when(
      Distance_from_Home == "Near" ~ 1,
      Distance_from_Home == "Moderate" ~ 2,
      Distance_from_Home == "Far" ~ 3
    ),
    Extracurricular_Activities = ifelse(Extracurricular_Activities == "Yes", 1, 0),
    Internet_Access = ifelse(Internet_Access == "Yes", 1, 0),
    Learning_Disabilities = ifelse(Learning_Disabilities == "Yes", 1, 0)
  )
```

## Box Plot
```{r}
# Assuming df exists and contains the Exam_Score column
boxplot(df$Exam_Score,
        main = "Box Plot of Exam Scores",
        ylab = "Exam Score",
        col = "lightblue",
        border = "blue")
```

## Remove Outlier
```{r}
# Calculate Q1, Q3, and IQR
Q1 <- quantile(df$Exam_Score, 0.25, na.rm = TRUE)
Q3 <- quantile(df$Exam_Score, 0.75, na.rm = TRUE)
IQR <- Q3 - Q1

# Define lower and upper bounds
lower_bound <- Q1 - 1.5 * IQR
upper_bound <- Q3 + 1.5 * IQR

# Filter the data to exclude outliers
df <- df[df$Exam_Score >= lower_bound & df$Exam_Score <= upper_bound, ]
```

## Box Plot After Outlier Removal
```{r}
# Assuming df exists and contains the Exam_Score column
boxplot(df$Exam_Score,
        main = "Box Plot of Exam Scores",
        ylab = "Exam Score",
        col = "lightblue",
        border = "blue")
```

## Summary Statistics
```{r}
summary(df)
```

## 2-d Scatter Plots
```{r}
# Save the plots to a single file
png("pairplots_after.png", width = 720, height = 720)

# Setting up a multi-panel plot layout
par(mfrow = c(ceiling(sqrt(ncol(df) - 1)), ceiling(sqrt(ncol(df) - 1))),
    mar = c(4, 4, 2, 1))

# Loop through all predictors (excluding Exam_Score)
for (col in names(df)) {
  if (col != "Exam_Score" && is.numeric(df[[col]])) {
    plot(
      df[[col]], df$Exam_Score,
      xlab = col, ylab = "Exam_Score",
      pch = 16, col = "blue"
    )
  }
}

dev.off()
```

## Heatmap
Values closer to 0 implies no correlation, values closer to 1 implies positive, values closer to -1 implies negative correlation.

```{r}
png("heat_map.png", width = 720, height = 720)

# Compute correlation matrix
cor_matrix <- cor(df[, sapply(df, is.numeric)], use = "complete.obs")

# Convert the correlation matrix into a data frame for ggplot2
cor_df <- as.data.frame(as.table(cor_matrix))

# Plot heatmap with correlation values
ggplot(cor_df, aes(Var1, Var2, fill = Freq)) +
  geom_tile(color = "white") +  # Create the heatmap with white gridlines
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0,
                       limit = c(-1, 1), name = "Correlation") +
  geom_text(aes(label = round(Freq, 2)), color = "black", size = 4) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Correlation Heatmap", x = "", y = "")

dev.off()
```

## Plotting Qualitative Data
```{r}
# Identify qualitative variables
qualitative_vars <- names(df)[sapply(df, is.factor) | sapply(df, is.character)]

# Generate a box plot for each qualitative variable
for (var in qualitative_vars) {
  print(
    ggplot(df, aes_string(x = var, y = "Exam_Score")) +
      geom_boxplot(fill = "lightblue", color = "darkblue") +
      labs(title = paste("Box Plot of Exam_Score by", var),
           x = var,
           y = "Exam_Score") +
      theme_minimal() +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))
  )
}
```

# 4 Linear Model
```{r}
# Step 1: Create the initial linear model
model <- lm(Exam_Score ~ ., data = df)

# Step 2: Check the summary for statistical significance
summary(model)
```

### Normality Check
```{r}
# Anderson-Darling Test for Residuals
ad.test(model$residuals)

# Visualize Residuals
# Histogram of residuals
ggplot(data.frame(residuals = model$residuals), aes(x = residuals)) +
  geom_histogram(aes(y = ..density..), bins = 30, fill = "blue", alpha = 0.5) +
  geom_density(color = "red", size = 1) +
  ggtitle("Histogram of Residuals")

# Q-Q Plot
qqnorm(model$residuals)
qqline(model$residuals, col = "red")
```

### Constant Variance Check
```{r}
# Breusch-Pagan Test
bptest(model)  # Null hypothesis: Homoscedasticity

# Residuals vs. Fitted Plot
plot(model$fitted.values, model$residuals,
     xlab = "Fitted Values", ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red")
```

### Remove insignificant parameters
Insignificant parameters are Gender, School_Type

```{r}
df <- df[, !names(df) %in% c("School_Type", "Gender")]
head(df)
```

```{r}
# Step 1: Create the initial linear model
model <- lm(Exam_Score ~ ., data = df)

# Step 2: Check the summary for statistical significance
summary(model)

# Step 3: Get R^2
summary(model)$r.squared
```

### Model with the 6 highest correlated variables in heatmap
```{r}
# Step 1: Create the initial linear model
model <- lm(Exam_Score ~ Hours_Studied + Attendance
            + Parental_Involvement + Access_to_Resources
            + Previous_Scores + Tutoring_Sessions, data = df)

# Step 2: Check the summary for statistical significance
summary(model)

# Step 3: Get R^2
summary(model)$r.squared
```